---
title: 'COURSERA - Reproducible Research: Peer Assessment 1'
author: "Stanislav Gerasymenko"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data

Load required packages:
```{r message = FALSE}
require(dplyr)
require(ggplot2)
require(gridExtra)
```

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 3.5, fig.path='Figures/', 
                      message = FALSE)
```

Set the working directory (**should be customised by the user**) and read the
data into R:
```{r}
wd <- paste("C:/!SG/OneDrive/Projects/SCIENCE/Coursera",
            "Coursera_05_Reproducible_Research/Course_Project1",
            "RepData_PeerAssessment1", sep = "/")
setwd(paste(wd, "INPUT", sep = "/"))
inputdf <- 
        read.csv("activity.csv")%>%
        mutate(interval = as.factor(interval))
# head(inputdf)
# str(inputdf)
# tail(inputdf)

```


## What is mean total number of steps taken per day?

Calculate the total steps walked during each day and make a histogram:
```{r}
totalStepsByDay <-
        inputdf %>%
        group_by(date) %>%
        summarise(steps = sum(steps))
# totalStepsByDay
```

```{r echo = FALSE}
plot1 <- ggplot(totalStepsByDay, aes(x = steps)) + 
        geom_histogram(alpha = 0.8, stat = "bin", binwidth = 1000) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme(text = element_text(size = 10))
plot1
```

Median value of steps walked during a day is:
```{r}
totalStepsByDay %>%
        summarise(steps = median(steps, na.rm = TRUE))
```

Mean value of steps walked during a day is:
```{r}
totalStepsByDay %>%
        summarise(steps = mean(steps, na.rm = TRUE))
```


## What is the average daily activity pattern?

Calculate and plot the mean steps walked during each interval over the 
whole period:
```{r}
totalStepsByInterval <-
        inputdf %>%
        mutate(interval = as.factor(interval)) %>%
        group_by(interval) %>%
        summarise(steps = mean(steps, na.rm = TRUE))
# str(totalStepsByInterval)
# tail(totalStepsByInterval)

```

```{r echo = FALSE}
plot2 <- ggplot(totalStepsByInterval, 
                aes(x = interval, y = steps, group = "identity")) + 
        geom_area(alpha = 0.8) + 
        scale_x_discrete(breaks = seq(0, 2400, 100)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme(text = element_text(size = 10))
plot2

```

Find the interval with maximum steps walked on average:
```{r}
totalStepsByInterval %>%
        filter(steps == max(steps))

```

## Imputing missing values

Calculate the number of missing values before imputing the missing values:
```{r}
inputdf %>%
        filter(is.na(steps)) %>%
        summarise(count = n())

```

Impute the missing values with average values for the corresponding interval:
```{r}
temp <- left_join(inputdf, totalStepsByInterval, by = "interval")
inputdfCorrected <- inputdf
inputdfCorrected[is.na(inputdf$steps), "steps"] <- 
        temp[is.na(inputdf$steps), "steps.y"]
rm(temp)

```

Calculate the number of missing values after imputing missing values:
```{r}
inputdfCorrected %>%
        filter(is.na(steps)) %>%
        summarise(count = n())
```

Calculate the total steps walked during each day after imputing the missing 
values:
```{r}
totalStepsByDayCorrected <-
        inputdfCorrected %>%
        group_by(date) %>%
        summarise(steps = sum(steps))
# totalStepsByDayCorrected
```

Median value of steps walked during a day after imputing the missing values is:
```{r}
totalStepsByDayCorrected %>%
        summarise(steps = median(steps, na.rm = TRUE))
```

Mean value of steps walked during a day after imputing the missing values is:
```{r}
totalStepsByDayCorrected %>%
        summarise(steps = mean(steps, na.rm = TRUE))
```

Calculate the mean steps walked during each interval over the whole period after
imputing missing values:
```{r}
totalStepsByIntervalCorrected <-
        inputdf %>%
        mutate(interval = as.factor(interval)) %>%
        group_by(interval) %>%
        summarise(steps = mean(steps, na.rm = TRUE))
# str(totalStepsByIntervalCorrected)
# tail(totalStepsByIntervalCorrected)

```

Join original and imputed tables for convenient graphing and see the
comparisons on the plots:
```{r}
totalStepsByDayComparison <- rbind(totalStepsByDay, totalStepsByDayCorrected)
totalStepsByDayComparison$Imputed <- 
        rep(c("Original", "Imputed"), 
            times = c(nrow(totalStepsByDay), nrow(totalStepsByDayCorrected)))
totalStepsByDayComparison$Imputed <- 
        as.factor(totalStepsByDayComparison$Imputed)

totalStepsByIntervalComparison <- 
        rbind(totalStepsByInterval, totalStepsByIntervalCorrected)
totalStepsByIntervalComparison$Imputed <- rep(c("Original", "Imputed"),
                           times = c(nrow(totalStepsByInterval),
                                     nrow(totalStepsByIntervalCorrected)))
totalStepsByIntervalComparison$Imputed <- 
        as.factor(totalStepsByIntervalComparison$Imputed)

```

Histograms of steps walked during each day before and after imputing the missing
values:  
```{r echo = FALSE}
plot3 <- ggplot(totalStepsByDayComparison, aes(x = steps, fill = Imputed)) + 
        geom_histogram(alpha = 0.8, stat = "bin", binwidth = 1000, 
                       position = "dodge") +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme(text = element_text(size = 10))
plot3

```

Mean steps walked during each interval over the whole period before and after 
imputing the missing values:          
```{r echo = FALSE}
plot4 <- ggplot(totalStepsByIntervalComparison, 
                aes(x = interval, y = steps, group = Imputed, fill = Imputed)) + 
        geom_area(alpha = 0.5, position = "dodge") + 
        scale_x_discrete(breaks = seq(0, 2400, 100)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme(text = element_text(size = 10))
plot4

```

## Are there differences in activity patterns between weekdays and weekends?

Create a factor with values WorkingDay/Weekend:
```{r}
Sys.setlocale("LC_TIME", "C")

inputdfCorrected <-
        inputdf %>%
        mutate(weekday = weekdays(as.Date(date))) %>%
        mutate(weekday = ifelse(weekday == "Monday" | 
                                        weekday ==  "Tuesday" |
                                        weekday ==  "Wednesday" |
                                        weekday ==  "Thursday" | 
                                        weekday ==  "Friday", 
                                "WorkingDay", "Weekend")) %>%
        mutate(weekday = as.factor(weekday))
# str(inputdfCorrected)

```

Calculate and plot the mean steps walked during each interval 
over WorkingDay/ Weekend periods after imputing missing values:
```{r}
totalStepsByIntervalWeekdayCorrected <-
        inputdfCorrected %>%
        mutate(interval = as.factor(interval)) %>%
        group_by(interval, weekday) %>%
        summarise(steps = mean(steps, na.rm = TRUE))
# str(totalStepsByInterval)
# tail(totalStepsByInterval)

```

```{r echo = FALSE}
plot5 <- ggplot(totalStepsByIntervalWeekdayCorrected, 
                aes(x = interval, y = steps, fill = weekday, group = weekday)) + 
        geom_area(alpha = 0.6, position = "dodge") + 
        scale_x_discrete(breaks = seq(0, 2400, 100)) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        theme(text = element_text(size = 10))
plot5

```

```{r echo = FALSE}
setwd(wd)
```